Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-12-06 analysis/paper/paper.Rmd
Updated xrf violin plot

```{r load-packages-and-data}
library(ChemoSpec)
library(data.table)
library(factoextra)
library(ggbiplot)
library(ggplot2)
library(ggsci)
library(grid)
library(gridBase)
library(gridExtra)
library(missMDA)
library(patchwork)
library(prospectr)
library(tidyverse)
library(spectrolab)

@@

# Add column with elemental groups for visualisation
Points_long <-
  Points_long %>%
    mutate(Group = case_when(
      endsWith(Element, "Mg") ~ "Alkaline Earth",
      endsWith(Element, "Al") ~ "Basic metal",
      endsWith(Element, "Si") ~ "Semimetal",
      endsWith(Element, "P") ~ "Nonmetal",
      endsWith(Element, "S") ~ "Nonmetal",
      endsWith(Element, "K") ~ "Alkali Metal",
      endsWith(Element, "Ca") ~ "Alkaline Earth",
      endsWith(Element, "V") ~ "Transition Metal",
      endsWith(Element, "Mn") ~ "Transition Metal",
      endsWith(Element, "Fe") ~ "Transition Metal",
      endsWith(Element, "Zn") ~ "Transition Metal",
      endsWith(Element, "Sr") ~ "Alkaline Earth",
      endsWith(Element, "Y") ~ "Transition Metal",
      endsWith(Element, "Zr") ~ "Transition Metal",
      endsWith(Element, "Ba") ~ "Alkaline Earth",
      endsWith(Element, "Ti") ~ "Transition Metal",
    ))

#Calculate and add value equal to 95% of the value range for each element, this is in preparation for 
#text annotations in final plot where a text will be positioned at this values height in each plot
Points_long <- 
  Points_long %>% 
  group_by(Element) %>% 
    mutate(ypos = min(Value)+0.95*diff(range(Value)))

#Prepare colors
loading.colors <- c("darkorange", "darkolivegreen3", "skyblue1", "lightslateblue", "salmon1", "lightgoldenrod1")

@@

#Violin plot of elemental content, trimmed, with fill based on elemental groups and number of values under LOD as text annotation
Points_long %>%
    ggplot(aes(x=Element, y=Value, fill = Group)) +
    geom_violin(trim = TRUE) +
    facet_wrap( ~ Element, scales = "free", labeller = labeller(Element = Elements)) +
    scale_fill_manual(values=loading.colors) +
    theme_bw() +
    geom_text(data=na_values, aes(x=0.2, y=ypos, label=na_values, fontface = "bold"), nudge_x = 0.4, inherit.aes = FALSE) +
    theme(
      legend.title = element_text(size = 12, face = "bold", colour = "black"),
      strip.text.x = element_text(size = 12, face = "bold", colour = "black"),
      axis.title.x = element_blank(),
      plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
      axis.title.y=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank())
```

>>>>

```{r load-packages-and-data}
library(ChemoSpec)
library(data.table)
library(factoextra)
library(ggbiplot)
library(ggplot2)
library(ggh4x)
library(ggsci)
library(grid)
library(gridBase)
library(gridExtra)
library(missMDA)
library(patchwork)
library(prospectr)
library(tidyverse)
library(spectrolab)

@@

# Add column with elemental groups for visualisation
Points_long <-
  Points_long %>%
    mutate(Group = case_when(
      endsWith(Element, "Mg") ~ "Lithophile",
      endsWith(Element, "Al") ~ "Lithophile",
      endsWith(Element, "Si") ~ "Lithophile",
      endsWith(Element, "P") ~ "Lithophile",
      endsWith(Element, "S") ~ "Chalcophile",
      endsWith(Element, "K") ~ "Lithophile",
      endsWith(Element, "Ca") ~ "Lithophile",
      endsWith(Element, "V") ~ "Lithophile",
      endsWith(Element, "Mn") ~ "Siderophile",
      endsWith(Element, "Fe") ~ "Siderophile",
      endsWith(Element, "Zn") ~ "Chalcophile",
      endsWith(Element, "Sr") ~ "Lithophile",
      endsWith(Element, "Y") ~ "Lithophile",
      endsWith(Element, "Zr") ~ "Lithophile",
      endsWith(Element, "Ba") ~ "Lithophile",
      endsWith(Element, "Ti") ~ "Chalcophile",
    ))

#Calculate and add value equal to 90% of the value range for each element, this is in preparation for 
#text annotations in final plot where a text will be positioned at this values height in each plot
Points_long <- 
  Points_long %>% 
    group_by(Element) %>% 
    dplyr::mutate(ypos = quantile(Value, 0.95))

#Prepare colors
loading.colors <- c("lightgoldenrod1", "darkorange", "lightpink1")

@@

#Prepare ylimits for each group in order to exclude extreme outliers in the violin plot graphic
ylims <- Points_long %>%
  group_by(Element) %>% 
  dplyr::summarise(y1 = 0, y2 = quantile(Value, 0.95))

#Prepare a list with formula for setting new ylimits for each group in ggh4x package "facetted_pos_scales() which makes it possible to work with facet_wrap()
scales <- c(
  Element == '12Mg' ~ scale_y_continuous(limits = c(as.vector(t(ylims[1,2:3])))),
  Element == '13Al' ~ scale_y_continuous(limits = c(as.vector(t(ylims[2,2:3])))),
  Element == '14Si' ~ scale_y_continuous(limits = c(as.vector(t(ylims[3,2:3])))),
  Element == '15P' ~ scale_y_continuous(limits = c(as.vector(t(ylims[4,2:3])))),
  Element == '16S' ~ scale_y_continuous(limits = c(as.vector(t(ylims[5,2:3])))),
  Element == '19K' ~ scale_y_continuous(limits = c(as.vector(t(ylims[6,2:3])))),
  Element == '20Ca' ~ scale_y_continuous(limits = c(as.vector(t(ylims[7,2:3])))),
  Element == '23V' ~ scale_y_continuous(limits = c(as.vector(t(ylims[8,2:3])))),
  Element == '25Mn' ~ scale_y_continuous(limits = c(as.vector(t(ylims[9,2:3])))),
  Element == '26Fe' ~ scale_y_continuous(limits = c(as.vector(t(ylims[10,2:3])))),
  Element == '30Zn' ~ scale_y_continuous(limits = c(as.vector(t(ylims[11,2:3])))),
  Element == '38Sr' ~ scale_y_continuous(limits = c(as.vector(t(ylims[12,2:3])))),
  Element == '39Y' ~ scale_y_continuous(limits = c(as.vector(t(ylims[13,2:3])))),
  Element == '40Zr' ~ scale_y_continuous(limits = c(as.vector(t(ylims[14,2:3])))),
  Element == '56Ba' ~ scale_y_continuous(limits = c(as.vector(t(ylims[15,2:3])))),
  Element == '81Ti' ~ scale_y_continuous(limits = c(as.vector(t(ylims[16,2:3]))))
)

#Violin plot of elemental content, trimmed, with fill based on elemental groups and number of values under LOD as text annotation
Points_long %>%
    ggplot(aes(x=Element, y=Value, fill = Group)) +
    geom_violin(trim = TRUE) +
    facet_wrap( ~ Element, scales = "free", labeller = labeller(Element = Elements)) +
    scale_fill_manual(values=loading.colors) +
    facetted_pos_scales(y = scales) +
    theme_bw() +
    geom_text(data=na_values, aes(x=0.2, y=ypos, label=na_values, fontface = "bold"), nudge_x = 0.4, inherit.aes = FALSE) +
    theme(
      legend.title = element_text(size = 12, face = "bold", colour = "black"),
      strip.text.x = element_text(size = 12, face = "bold", colour = "black"),
      axis.title.x = element_blank(),
      plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
      axis.title.y=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank())
```
